---
title: "Pseudo bulk chondrosarcoma analysis"
author: "Léa ROGUE"
output:
  pdf_document:
    fig_caption: true
    latex_engine: xelatex
    toc: true
    toc_depth: '4'
    includes:
      before_body: ../tex_files/list_of_figures.tex
date: "2025-03-25"
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  tidy.opts = list(width.cutoff = 60), 
  tidy = TRUE, 
  fig.align = 'center'
)
```

\newpage

This script analyze pseudo bulk scRNAseq the same manner the microarray part to see if it is coherent and to see the infiltrated tumors or not.

# Load libraries
```{r, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ComplexHeatmap)
library(colorRamp2)
```

# Read data
```{r}
df_tpm <- read.table("../../results/matrix_pseudo_bulk_tpm_normalized.tsv", sep = "\t", header = TRUE, check.names = F)

# Read CTA file
df_CTA <- read.table("../../data/CTA_list_clean.txt", header = FALSE)

# Read immune cells
df_immune_sign <- read.table("../../data/immune_cells_mhc_genes.tsv", header = FALSE, sep = "\t")

# Read list of CTA that impact survival probabilities
l_CTA_conv <- read.table("../../data/CTA_signif_coxph_conv_indiv.txt", sep = "\t", header = FALSE)$V1
```

# Prepare data
```{r}
df_tpm$CTA <-  NA

# Rename col
colnames(df_CTA) <- c("SYMBOL")

# Df with CTA and whole genes by adding CTA in col CTA for CTA genes
df_tpm_CTA <- df_tpm %>%
  left_join(df_CTA, by = "SYMBOL") %>%
  mutate(CTA = ifelse(SYMBOL %in% df_CTA$SYMBOL, "CTA", NA))

# Reorganize columns
df_tpm_CTA <- df_tpm_CTA %>%
  select(SYMBOL, CTA, everything())

# Read immune cells genes
colnames(df_immune_sign) <- c("Signature", "Gene")

# Merge df_CTA_whole and df_immune_sign to associate signatures
df_tpm_CTA_imm_cells <- merge(df_tpm_CTA, df_immune_sign, by.x = "SYMBOL", by.y = "Gene", all.x = TRUE)

# Reorganize the columns if needed
df_tpm_CTA_imm_cells <- df_tpm_CTA_imm_cells %>%
  select(SYMBOL, CTA, Signature, everything())

# Combine multiple signatures into one for each gene by concatenating with a comma
df_tpm_CTA_imm_cells_clean <- df_tpm_CTA_imm_cells %>%
  group_by(SYMBOL) %>%
  summarise(
    Signature = paste(unique(Signature), collapse = ", "),                                              
    across(everything(), ~first(.)),  
    .groups = "drop")

# Average the expression between same immune cells types
# Take rows with immune cells signature from normalized data
df_avg_immune_sign <- df_tpm_CTA_imm_cells %>%
  filter(Signature != "NA")

# Group by signature and calculate mean of expression values
df_avg_immune_sign_final <- df_avg_immune_sign %>%
  select(-c(SYMBOL, CTA)) %>% 
  group_by(Signature) %>%
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE))) 
```


# I. Relative immune cells expression
```{r, fig.cap="Relative immune cells expression heatmap"}
# Create the heatmap
heatmap_data <- as.data.frame(df_avg_immune_sign_final)
rownames(heatmap_data) <- heatmap_data$Signature
heatmap_data <- heatmap_data[ , -1]  # Remove the Signature column
heatmap_data <- log2(heatmap_data+1)

# Calculate Z-scores
z_scores_row <- t(scale(t(heatmap_data)))

# Add columns
df_z_scores <- as.data.frame(z_scores_row)
colnames(df_z_scores) <- c("Low_L07", "Low_L28", "High_L31", "High_L44", "Ben_L49", "Med_L63", "Med_L80", "Low_L81", "Ded_L83")
  
colors <- colorRampPalette(c("blue", "white", "red"))(100)
Heatmap(
    as.matrix(df_z_scores),
    cluster_rows = TRUE,    
    cluster_columns = TRUE,
    cluster_column_slices = TRUE,
    clustering_distance_columns = "euclidean",
    clustering_method_columns = "complete",
    show_column_dend = TRUE,
    col = colorRamp2(seq(-8, 8, length.out = 100), colors),  
    border = NA,             
    show_column_names = TRUE,
    column_names_gp = gpar(fontsize = 7),
    row_names_gp = gpar(fontsize = 7),
    heatmap_legend_param = list(title = "Expression Level")
)

```
We see that the more infiltrated tumors are dedifferentiated and low tumors, contrary to the less infiltrated that are high grade.

# II. Relative expression of CTAs
```{r, fig.height=12, fig.cap="Relative expression of CTAs"}
# Create the heatmap
cta_data <- subset(df_tpm_CTA, df_tpm_CTA$CTA == "CTA")
rownames(cta_data) <- cta_data$SYMBOL
cta_data <- cta_data[ , -c(1, 2)]
cta_data <- log2(cta_data+1)

# Calculate Z-scores
z_scores_row <- t(scale(t(cta_data)))

# Add columns
df_z_scores <- as.data.frame(z_scores_row)

Heatmap(
    as.matrix(df_z_scores),
    cluster_rows = TRUE,    
    cluster_columns = TRUE,
    cluster_column_slices = TRUE,
    clustering_distance_columns = "euclidean",
    clustering_method_columns = "complete",
    show_column_dend = TRUE,
    col = colorRamp2(seq(-8, 8, length.out = 100), colors),  
    border = NA,             
    show_column_names = TRUE,
    column_names_gp = gpar(fontsize = 7),
    row_names_gp = gpar(fontsize = 4),
    heatmap_legend_param = list(title = "Expression Level")
)
```

